Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-04-16 versions/response.v1.txt

> Associate Editor
> Editors’ comments to the author:
> All four reviewers agreed that the work is worth publishing and
> provides an important insight into the behavior of the joint
> ML/ancestral states estimation method. However, all reviewers asked
> for what amounts to major revisions to the text and presentation. You
> should especially focus on clarity and accessibility, provide the
> requested examples, and address suggestions on how the results could
> be even further improved.

> Finally, please pay VERY careful attention to improving your
> literature review (reviewer 4) and making a very clear case about how
> your work IMPROVES and EXTENDS previous results. And, please, do get
> rid of the Farris zone. The name "Farris" should not be attached to
> results that have to do with ML estimation, given the long and
> torturous efforts of cladists to undermine the entire line of inquiry.

Felsenstein's book has a subsection entitled "The Farris zone" spanning two pages (pp 120-121), a figure entitled "A tree in the Farris zone.", and a table with a caption "Properties of the Felsenstein and Farris zones..."
The term "Farris zone" has almost 7 times as many citations as the term "inverse Felsenstein zone" on Google Scholar.
Thus, although we realize that the term causes offense, it is standard.

> Reviewer: 1

> Comments to the Author
> The authors show that when ancestral states are considered as
> parameters, there are zones where topologies and edge-lengths are
> inconsistently estimated. As is evidenced by any ML package, it is
> widely accepted that ancestral states should not be parameters; the
> large number of parameters suggests the approach will be problematic.
> Still, to the best of my knowledge, it has not been shown that
> treating ancestral states as parameters actually leads to poor
> performance, or how poor. Moreover, the authors point to Sagulenko et
> al (2018) - now published in Virus Evolution - as an example
> implementation. I found the paper interesting and I believe a more
> succinct presentation will be of interest.

> A main concern with the paper is that the shaded regions of
> topological inconsistency or poor estimation in Figs 2-4, are only
> subsets of the actual regions where topological inconsistency or poor
> estimation occur; and yet determining the actual regions seems
> feasible. The contrast between Fig 5 and 4 illustrates how far off the
> subsets are in giving the actual regions where estimation is
> problematic. The white region in Fig 5 is the actual region where hat
> w=1 whereas the shaded region of Fig 3 is the subset of the actual
> region that the authors were able to determine through analytical
> arguments. For most readers, out of Fig 3 and Fig 5, only Fig 5 is of
> interest.

> Getting exact regions is what I think the authors should do. The
> approach used to obtain Fig 5 is generally applicable. Explicit,
> relatively simple, expressions are given in the appendix for long-run
> lnls, l_{tau_1,t*}(tau_2,w,x,y). For any given (x*,y*) on a fine grid
> in Figs 2-4, l_{tau_1,t*}(tau_j,w,x,y), can be maximized using
> numerical algorithms over the three parameters w,x,y. Doing so for
> tau_j=tau_1, tau_2 would determine the actual regions of problematic
> estimation corresponding to Figs 2-4. The only potential difficulty
> that I can see is that numerical algorithms might not converge. This
> can either be avoided by considering a cubic grid of starting (w,x,y),
> or through something like the following. For any given set of
> ancestral character state assignments for the distinct set of 8
> patterns 1111, 1110,... (0000 can treated as the same as 1111), the
> long-run lnl is of the form

> l(x,y,w)=a log(1-x) + b log(1+x) + c log(1-y) + d log(1+y) + e log(1-w) + f log(1+w)

> where a,..f>=0 are constants determined by (x*,y*) and what ancestral
> states have been assigned. The function l(x,y,w) has an explicit
> simple maximizer. Thus the overall maximizer is determined as the
> largest maximized l(x,y,w) over all 4^8 possible ancestral state
> assignments: a finite number of comparisons. The actual number of
> possible ancestral state assignments is quite a bit smaller than 4^8
> (for instance 11 is always the ancestral reconstruction for 1111) and
> the authors have a lot of familiarity with what the ancestral
> reconstructions are. I believe Fig 2 can be replaced with an exact (up
> to grid approximation) region of topological inconsistency and Figs
> 3-5 with heat maps of limiting (x,y,w) for tau_1.

Theory leads to generalization.

> The other main concern that I have is with the restriction that
> t=(x,y,x,y,w) in estimation rather than t=(t_1,...,t_5). On the one
> hand, usually one expects that incorporating knowledge about the true
> underlying tree (t_1=t_3 & t_2=t_4 in the present case) should be
> helpful for a good estimation procedure. Thus the fact that treating
> ancestral states as parameters gives inconsistent topological
> estimation in this setting demonstrates that the approach is
> problematic. On the other hand, allowing t=(t_1,...,t_5) is what would
> be done in any practical implementation and it seems quite possible to
> me that results would be radically different in this case. For
> instance, from Fig 2, the generating tree with all equal edge-lengths
> (y*=x*=0.2) provides a case where topology tau_2 is inconsistently
> estimated with the restriction that estimated t=(x,y,x,y,w). Because
> of the symmetry of the problem, I am doubtful y*=x*=0.2 would be a
> point of inconsistency if t=(t_1,...,t_5) were unrestricted in
> estimation. The issue deserves discussion because I don't think it is
> transparent.

> I don't think the restricted case of p8-9 will be of interest to
> readers. Only the general case.

> Minor Comments:

> -l78-88: Calculations are more natural with `fidelities' and this
> might remain in the appendix but it is much more conventional in the
> literature for x,y,w to be edge-lengths or probabilities of a
> different character states at the tips of the edge. It takes quite a
> bit of getting used to realize that (0,0) corresponds to infinite
> edge-lengths and it seems unnecessary with a little work on the part
> of the authors to do the conversions that will make the figs more
> natural for readers.

> Reviewer: 2

> Comments to the Author
> In this manuscript  the authors show that joint inference of phylogeny
> and ancestral characters is not consistent in general by providing a
> counter-example, in which as the number of sites grows, they show that
> the wrong tree topology can have a higher likelihood than the correct
> topology under the joint inference model.
> The result is important for the community to understand, in particular
> because TreeTime (the joint inference method) is used in high-profile
> analyses on Ebola, influenza and other viruses on the nextflu /
> nextstrain websites. These are beautiful and accessible sites aiming
> to make analyses on viruses open to exploration by the community, but
> this comes at the cost of suppressing the details of the analysis. If
> indeed there are deep statistical issues or limitations it is
> important that the community knows what these are and what their
> implications are.

> Accordingly my comments are focussed on where I think the manuscript
> could be more clear (and where I did not quite follow the logic). My
> major suggestion would be to go even further towards demonstrating in
> a simple example how the joint inference can be wrong – for example,
> providing a toy multiple sequence alignment or even one of some
> influenza HA sequences – you could find 2 sets of 4 closely-related
> sequences that have the Farris and Felsenstein topologies, maybe? And
> demonstrate that the inference gives the wrong result by explicitly
> computing the likelihood.

> Also, laying out the theory a little bit more step-by-step so that a
> general quantitative reader doesn't struggle too much to follow would
> help (suggestions below).

> I know that it's not feasible to explicitly compute (analytically)
> these likelihoods for bigger trees. But one of the main motivations
> and applications for tools like TreeTime is surely that the number of
> sequences grows (not the number of sites, which is fixed by the viral
> sequence length in the case of viruses). What happens when the number
> of tips grows? Are there more opportunities for the kind of issue that
> gives the wrong value for w? Many influenza trees have many 0 (or near
> 0) internal branch lengths, and it would be interesting to understand
> the implications for the ancestral state reconstructions (which in
> turn are used to infer spatial spread, transmission etc).

> Some detailed comments and questions:
> – line 85 some cut/paste or sentence issue
> - line 97 below (2) define A^p
> - l106: why is the functional form of (3) determined by inequalites;
> what are they?
> - l 118-124: clarify are tau1 and tau2 Farris and Felsenstein?
> L 132 clarify “equivalent but different notion” – different from what?
> - line 146 “seeing no change along a branch more likely increases the
> likelihood by introducing a term of (1 + θ) instead of (1 − θ)” – I
> don't understand this
> line 185++ and figure5: true value of w?

> L 334 missing “the”

> L332++ I got confused about the complements and factors of 2, though I
> see the point that the site split has symmetry. As it's fundamental to
> the rest of the work, a more clear statement of where the 2 comes from
> in the second eq below 358 would be very helpful.

> L354 why does psi take the complement of site patterns? Just above it
> is defined as a random variable psi (Y) where Y is a site pattern

> What are the elements of the sets eta(tau, t)?

> Why is xi surjective? I don't understand the equation before line 359,
> and then I don't understand what “such a choice” is in line 359
> line 364 define p(e)

> line 366  I don't understand (8) but perhaps that's just background I
> don't have – a quick explanation or reference would be handy
> What's a 'generating probability' ? Is that the fidelity? Or a
> generator?

> Reviewer: 3

> Comments to the Author
> Shaw and Matsen show that a joint maximum-likelihood inference of
> phylogenies and ancestral states is not consistent when branch lengths
> are long. We have recently proposed such a joint estimation for
> sequences that recently diverged and are hence similar (Sagulenko et
> al). The results presented here are anticipated in Sagulenko et al
> "For more divergent sequences, however, subleading states of internal
> nodes make a substantial contribution and the iterative optimization
> will underestimate the branch lengths."
> and
> "The iterative optimization procedures are not expected to be accurate
> for trees were many sites are saturated. In scenarios with extensive
> uncertainty of ancestral states and tree topology, convergence of the
> iterative steps cannot be guaranteed."

> In this light, it should be pointed out that joint estimation
> procedure was explicitly suggested as an approximation and hence never
> expected to be consistent in a strict sense. Furthermore, it was
> suggested as an approximation when branch lengths are short -- the
> opposite of the problematic regions identified here.
> But systematic analysis of the behavior of this approximation is
> useful and I believe the work is correct (but haven't checked the
> proofs). The work by Shaw and Matsen suggests the radius of
> convergence of this procedure is somewhere around branch length of
> order 10-20%, which is consistent with results reported by Sagulenko
> et al.

> At the same time, these results are not particularly surprising.
> Assigning ML ancestral states will always result in shorter branches.
> For short branches, this correction is linear in branch length, for
> longer branches higher order effects start to contribute. If the
> typical branch length around a bifurcating node is x<<1, a fraction ~x
> of the sites mutated along one of the three adjacent branches, while a
> fraction ~x^2 mutated on two branches and not the third. The latter
> case is ignored in this joint inference procedure. Not having to sum
> over these subleading states is what makes joint inference faster, but
> such truncation will first lead to inaccuracies and at longer
> distances to qualitatively different solutions (like Taylor expanding
> 1/(1-x) ).

> There are other obvious cases where joint inference will fail. For
> sequences with a heavily biased in nucleotide composition (say biased
> towards A). For sufficiently long terminal branches, all ancestral
> states will be estimated as all 'A' and subsequent branch length
> estimate will shrink all internal branches to zero. This happens for
> the very same reason that the most likely outcome of multiple flips of
> a biased coin is all heads -- a very atypical yet most likely outcome.

> Currently, the work largely shows that areas of inconsistencies exist.
> It would be useful if the authors could present Taylor expansions for
> the ML branch length estimates around x=y=w=1. This would more
> explicitly show the inaccuracies accumulate with increasing branch
> length and how the inaccuracies depend on asymmetries in the tree. To
> me, this would be much more useful than to know that joint inference
> will fail if branches are long.

> Specific comments:
>  - when y=x=w, the two topologies are indistinguishable other than for
>  the labeling of the leaves. What happens in Fig 2 on the diagonal?
>  Does it degenerate into a star?

>  - line 50: overestimates branch lengths? It overestimates fidelity ->
>  underestimates branch length.  As expected.  Again line 249. Please
>  check. The back and forth between fidelity and branch length can be
>  confusing.

>  - Fig 5: it is surprising that \hat{w} seems completely flat as a
>  function of x* until it jumps discontinuously to 0. Why not include
>  the analytical bound in Fig 5?

>  - It would be nice if the abstract pointed out that these
>  inconsistencies arise when branch lengths are long and that the
>  approximate joint maximization procedure was suggested for precisely
>  the opposite limit. Similar at the beginning of the discussion, I
>  think it would be more appropriate if this would say smth along the
>  lines "Sagulenko et al suggested an approximation for short
>  branches... we show that this doesn't work for long branches..."

> Richard Neher

> Reviewer: 4

> Comments to the Author
> Review of Shaw, D. and Matsen, F.  "Joint maximum-likelihood of
> phylogenies and ancestral states is not consistent"

> Shaw and Matsen demonstrate that the approach of jointly estimating
> ancestral states and tree topologies (or branch lengths on a fixed
> topology) is not statistically consistent.  I believe that their
> results are correct, and that their result is important to
> disseminate.  The paper is clearly written, given the complexity of
> the material, and I have no reservation about recommending publication
> virtually as is.

> However, although my original intent was to carefully work through the
> math (including the derivations and proofs in the appendix), I have
> not found the time to do so, and because of the lateness of my review
> at this point, I will provide just a few comments in the hope that
> they will be useful.  Hopefully, other reviewers were able to check
> the correctness of the mathematical results more carefully.

> The only real criticism I have of the paper is that the authors could
> have done a better job in connecting their new work to previous
> results.  For example, Goldman (1990, Syst. Biol. 39:345-361) showed
> that joint estimation of topology and ancestral states by maximum
> likelihood was equivalent to maximum parsimony analysis for the same
> 2-state model considered by Shaw and Matsen when all branches are
> forced to have the same length.  They noted the statistical
> inconsistency of this approach, which was necessarily true because
> parsimony was already known to be inconsistent (the famous result of
> Felsenstein 1978).  Shaw and Matsen extend this result to the case of
> unequal branch lengths, which while not particularly surprising, is
> nonetheless an important extension.  The authors should also cite key
> relevant earlier work on the identifiability ot the tree topology with
> "classical" maximum likelihood phylogenetic estimation (Allman et al
> 2008, Adv Appl Prob 40:229–249; Chai and Housworth 2011, Syst Biol
> 60:713–718) in addition to crediting Roychoudhury 2015 for
> establishing ML consistency.

%EM Can you chat me links to these refs when you find them? I can get
%the pdfs for you as needed.

> Finally, let me just add that I hate the term "Farris Zone".  The
> "Felsenstein Zone" was a name applied to Felsenstein's important
> demonstration of a tree and branch-length configuration where
> parsimony would be an inconsistent estimator of the phylogeny.  On the
> contrary, Farris had nothing to do with the "Farris zone"; it was
> simply a name given by Mark Siddall to a different configuration in
> which parsimony outperforms likelihood, thinking incorrectly that
> likelihood was inconsistent in this zone.  Swofford et al. (2001,
> cited by the authors) referred to this zone as the "inverse
> Felsenstein zone", which I think is a better name for it.

### 2018-04-17 joint_inf.tex
fixing reference to Julie's grant

This work was supported by National Institutes of Health grants R01-GM113246, R01-AI12096, U19-AI117891, and U54-GM111274 as well as National Science Foundation grants CISE-1561334 and CISE-1564137.
The research of Frederick Matsen was supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation.

>>>>

This work was supported by National Institutes of Health grants R01-GM113246, R01-AI120961, U19-AI117891, and U54-GM111274 as well as National Science Foundation grants CISE-1561334 and CISE-1564137.
The research of Frederick Matsen was supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation.
